---
title: "Provider Change: Survival Analysis"
author: "Gulzira Igilmanova"
output:
  html_document:
    fig_width: 16
    fig_height: 9
    toc: true
    toc_depth: 4
params:
  inpatient_complete:    "output/survival_inpatient_complete.csv"
  inpatient_severe:      "output/survival_inpatient_severe.csv"
  inpatient_non_severe:  "output/survival_inpatient_non_severe.csv"
  daypatient_complete:   "output/survival_daypatient_complete.csv"
  daypatient_severe:     "output/survival_daypatient_severe.csv"
  daypatient_non_severe: "output/survival_daypatient_non_severe.csv"
---

## Dependencies and Setup

If you already don't have it, install the packages required to run the analysis.
```{r dependencies, eval = FALSE}
install.packages("kmi")
install.packages("mvna")
install.packages("etm")
install.packages("survival")
```

Now some setup is necessary:
```{r setup}
library(kmi)
library(mvna)
library(etm)
library(survival)

# read the CSV files
data_inpatient <- read.csv(params$inpatient_complete)
data_inpatient_sev <- read.csv(params$inpatient_severe)
data_inpatient_ns <- read.csv(params$inpatient_non_severe)

data_daypatient <- read.csv(params$daypatient_complete)
data_daypatient_sev <- read.csv(params$daypatient_severe)
data_daypatient_ns <- read.csv(params$daypatient_non_severe)

# aux variables
states <- c("0", "1", "2")
transitions <- c("0 1", "0 2", "1 2")
transition_matrix <- matrix(FALSE, 3, 3, dimnames = list(c(0, 1, 2), c(0, 1, 2)))
transition_matrix[1, 2:3] <- TRUE
transition_matrix[2, 3] <- TRUE

# Data Remap

# Inpatient
# sex remap W-1;M-0
data_inpatient$sex <- ifelse(data_inpatient$sex == "W", 1, 0)
data_inpatient_sev$sex <- ifelse(data_inpatient_sev$sex == "W", 1, 0)
data_inpatient_ns$sex <- ifelse(data_inpatient_ns$sex == "W", 1, 0)

# replace 99 with "cens" in the 'to' column
data_inpatient$to[data_inpatient$to == 99] <- "cens"
data_inpatient_sev$to[data_inpatient_sev$to == 99] <- "cens"
data_inpatient_ns$to[data_inpatient_ns$to == 99] <- "cens"

# Daypatient
# sex remap W-1;M-0
data_daypatient$sex <- ifelse(data_daypatient$sex == "W", 1, 0)
data_daypatient_sev$sex <- ifelse(data_daypatient_sev$sex == "W", 1, 0)
data_daypatient_ns$sex <- ifelse(data_daypatient_ns$sex == "W", 1, 0)

# replace 99 with "cens" in the 'to' column
data_daypatient$to[data_daypatient$to == 99] <- "cens"
data_daypatient_sev$to[data_daypatient_sev$to == 99] <- "cens"
data_daypatient_ns$to[data_daypatient_ns$to == 99] <- "cens"
```

## Tables

### Table 2. Table of observed transitions

```{r possible_transitions}
# Inpatient table of possible transitions
knitr::kable(table(data_inpatient$from, data_inpatient$to), caption = "Table 2.a. Inpatient Stay")

# Daypatient table of possible transitions
knitr::kable(table(data_daypatient$from, data_daypatient$to), caption = "Table 2.b. Daypatient Care")
```

### Table 3. Cox proportional hazards model (final State "Inpatient Stay")

```{r table3}
# INPATIENT STAY
# initial to provider change 01
cox_inpatient.01 <- coxph(
  Surv(entry, exit, to == 1) ~ as.factor(severe)
  + as.factor(comorbidity)
  + I(age / 10)
  + as.factor(sex),
  data_inpatient,
  subset = from == 0
)
summary(cox_inpatient.01)
cox.zph(cox_inpatient.01)

# initial to inpatient stay 02
cox_inpatient.02 <- coxph(
  Surv(entry, exit, to == 2) ~ as.factor(severe)
  + as.factor(comorbidity)
  + I(age / 10)
  + as.factor(sex),
  data_inpatient,
  subset = from == 0
)
summary(cox_inpatient.02)
cox.zph(cox_inpatient.02)

# from provider change to inpatient stay 12
cox_inpatient.12 <- coxph(
  Surv(entry, exit, to == 2) ~ as.factor(severe)
  + as.factor(comorbidity)
  + I(age / 10)
  + as.factor(sex),
  data_inpatient,
  subset = from == 1
)
summary(cox_inpatient.12)
cox.zph(cox_inpatient.12)
```

### Table 4. Cox proportional hazards model (final State "Daypatient Care")

```{r table4}
# DAYPATIENT CARE
# initial to provider change 01
cox_daypatient.01 <- coxph(
  Surv(entry, exit, to == 1) ~ as.factor(severe)
  + as.factor(comorbidity)
  + I(age / 10)
  + as.factor(sex),
  data_daypatient,
  subset = from == 0
)
summary(cox_daypatient.01)
cox.zph(cox_daypatient.01)

# initial to daypatient care 02
cox_daypatient.02 <- coxph(
  Surv(entry, exit, to == 2) ~ as.factor(severe)
  + as.factor(comorbidity)
  + I(age / 10)
  + as.factor(sex),
  data_daypatient,
  subset = from == 0
)
summary(cox_daypatient.02)
cox.zph(cox_daypatient.02)

# from provider change to daypatient care 12
cox_daypatient.12 <- coxph(
  Surv(entry, exit, to == 2) ~ as.factor(severe)
  + as.factor(comorbidity)
  + I(age / 10)
  + as.factor(sex),
  data_daypatient,
  subset = from == 1
)
summary(cox_daypatient.12)
cox.zph(cox_daypatient.12)
```

### Table 5. Cox proportional hazards of Transition 1 -\> 2 (Explorative Analysis of Covariates)

```{r table5}
# INPATIENT STAY
# from provider change to inpatient stay 12
summary(cox_inpatient.12)
cox.zph(cox_inpatient.12)

# DAYPATIENT CARE
summary(cox_daypatient.12)
cox.zph(cox_daypatient.12)
```

#### Markov Assumption tests

```{r markov_assumption, warning = FALSE}
# testing the Markov assumption for INPATIENT STAY
coxph(Surv(entry, exit, to == "2") ~ entry,
      data = subset(data_inpatient, from == "1"))

# testing the Markov assumption for DAYPATIENT CARE
coxph(Surv(entry, exit, to == "2") ~ entry,
      data = subset(data_daypatient, from == "1"))
```

### Table 6. Time dependent covariates

```{r table6}
# time-dep covariates for INPATIENT STAY
time_cox_inpatient <- coxph(
  Surv(entry, exit, to == 2) ~ as.factor(from) + as.factor(severe) +
    as.factor(comorbidity) + I(age / 10) +
    as.factor(sex),
  data_inpatient
)
summary(time_cox_inpatient)

# time-dep covariates for DAYPATIENT CARE
time_cox_daypatient <- coxph(
  Surv(entry, exit, to == 2) ~ as.factor(from) + as.factor(severe)
  + as.factor(comorbidity) + I(age / 10) +
    as.factor(sex),
  data_daypatient
)
summary(time_cox_daypatient)
```

## Figures

### Figure 2. Cumulative hazards and transition probabilities from `State - 1` to

`State - 2`

```{r figure2, results = 'hide'}
mvna.inpatient_sev <- mvna(data_inpatient_sev, states, transition_matrix, "cens")
mvna.daypatient_sev <- mvna(data_daypatient_sev, states, transition_matrix, "cens")
mvna.inpatient_ns <- mvna(data_inpatient_ns, states, transition_matrix, "cens")
mvna.daypatient_ns <- mvna(data_daypatient_ns, states, transition_matrix, "cens")

etm.inpatient_sev <- etm(
  data_inpatient_sev,
  states,
  transition_matrix,
  "cens",
  s = 0
)
etm.daypatient_sev <- etm(
  data_daypatient_sev,
  states,
  transition_matrix,
  "cens",
  s = 0
)
etm.inpatient_ns <- etm(
  data_inpatient_ns,
  states,
  transition_matrix,
  "cens",
  s = 0
)
etm.daypatient_ns <- etm(
  data_daypatient_ns,
  states,
  transition_matrix,
  "cens",
  s = 0
)

par(mfrow = c(1, 2))
plot(
  mvna.inpatient_sev,
  tr.choice = c("1 2"),
  col = "red",
  cex = 1.5,
  xlab = "",
  ylab = "",
  lwd = 1.8,
  legend = FALSE,
  cex.axis = 1.2,
  ylim = c(0, 1)
)
lines(
  mvna.inpatient_ns,
  tr.choice = c("1 2"),
  col = "green",
  cex = 1.5,
  lwd = 1.8
)
lines(
  mvna.daypatient_sev,
  tr.choice = c("1 2"),
  col = "blue",
  cex = 1.5,
  lwd = 1.8
)
lines(
  mvna.daypatient_ns,
  tr.choice = c("1 2"),
  col = "yellow",
  cex = 1.5,
  lwd = 1.8
)
par(family = "sans")
mtext(
  "Cumulative Hazard",
  side = 2,
  line = 3,
  font = 2,
  cex = 1.6
)
mtext(
  "Days",
  side = 1,
  line = 3,
  font = 2,
  cex = 1.6
)
legend(
  0.90,
  0.95,
  legend = c(
    'Inpatient:SMI',
    'Inpatient:non-SMI',
    'Daypatient:SMI',
    'Daypatient:non-SMI'
  ),
  col = c('red', 'green', 'blue', 'yellow'),
  lty = 1,
  cex = 1.2,
  y.intersp = 1,
  xjust = 0
)

#transition probability
plot(
  etm.inpatient_sev,
  tr.choice = c("1 2"),
  col = "red",
  cex = 1.5,
  xlab = "",
  ylab = "",
  lwd = 1.8,
  legend = FALSE,
  cex.axis = 1.2
)
lines(
  etm.inpatient_ns,
  tr.choice = c("1 2"),
  col = "green",
  cex = 1.5,
  lwd = 1.8
)
lines(
  etm.daypatient_sev,
  tr.choice = c("1 2"),
  col = "blue",
  cex = 1.5,
  lwd = 1.8
)
lines(
  etm.daypatient_ns,
  tr.choice = c("1 2"),
  col = "yellow",
  cex = 1.5,
  lwd = 1.8
)
par(family = "sans")
mtext(
  "Transition Probability",
  side = 2,
  line = 3,
  font = 2,
  cex = 1.6
)
mtext(
  "Days",
  side = 1,
  line = 3,
  font = 2,
  cex = 1.6
)
legend(
  0.90,
  0.95,
  legend = c(
    'Inpatient:SMI',
    'Inpatient:non-SMI',
    'Daypatient:SMI',
    'Daypatient:non-SMI'
  ),
  col = c('red', 'green', 'blue', 'yellow'),
  lty = 1,
  cex = 1.2,
  y.intersp = 1,
  xjust = 0
)
```

#### Confidence Intervals for Cumulative Hazards

```{r ci_ch}
summary(mvna.inpatient_sev)
summary(mvna.daypatient_sev)
summary(mvna.inpatient_ns)
summary(mvna.daypatient_ns)
```

#### Confidence Intervals for Transition Probabilities for SMI and non-SMI groups

```{r ci_tp}
summary(etm.inpatient_sev, tr.choice = c("1 2"))
summary(etm.daypatient_sev, tr.choice = c("1 2"))
summary(etm.inpatient_ns, tr.choice = c("1 2"))
summary(etm.daypatient_ns, tr.choice = c("1 2"))
```

### Figure 3. Cumulative hazards and transition probabilities of the complete model

```{r figure3, results = 'hide'}
# fulldata event: Inpatient stay
# nelson-aalen estimator of cumulative incidences
par(mfrow = c(2, 2), family = "sans")
line_colors <- c("darkgrey", "darkgreen", "darkred")

mvna.inpatient <- mvna(data_inpatient, states, transition_matrix, "cens")
summary(mvna.inpatient)

plot(
  mvna.inpatient,
  col = line_colors,
  main = "Event: Inpatient Stay",
  xlab = "Days",
  lty = 1,
  lwd = 2,
  cex = 1.2,
  legend = F
)
legend(
  "topleft",
  inset = c(0.05, 0.05),
  legend = transitions,
  col = line_colors,
  lty = 1,
  lwd = 2,
  cex = 1.2,
  bty = "o"
)

# aalen-johansen estimator of transition probabilities
etm.inpatient <- etm(
  data_inpatient,
  states,
  transition_matrix,
  "cens",
  s = 0,
  covariance = F
)
summary(etm.inpatient)

plot(
  etm.inpatient,
  tr.choice = transitions,
  col = line_colors,
  xlab = "Days",
  lty = 1,
  cex = 1.2,
  lwd = 2,
  font.lab = 2,
  legend = F
)
legend(
  "topleft",
  inset = c(0.05, 0.05),
  legend = transitions,
  col = line_colors,
  lty = 1,
  lwd = 2,
  cex = 1.2,
  bty = "o"
)

# fulldata event: Daypatient care
# nelson-aalen estimator of cumulative incidences
mvna.daypatient <- mvna(data_daypatient, states, transition_matrix, "cens")
summary(mvna.daypatient)

plot(
  mvna.daypatient,
  col = line_colors,
  main = "Event: Daypatient Care",
  xlab = "Days",
  lty = 1,
  lwd = 2,
  font.lab = 2,
  cex = 1.2,
  legend = F
)
legend(
  "topleft",
  inset = c(0.05, 0.05),
  legend = transitions,
  col = line_colors,
  lty = 1,
  lwd = 2,
  cex = 1.2,
  bty = "o"
)

# aalen-johansen estimator of transition probablities
etm.daypatient <- etm(
  data_daypatient,
  states,
  transition_matrix,
  "cens",
  s = 0,
  covariance = F
)
summary(etm.daypatient)

plot(
  etm.daypatient,
  tr.choice = transitions,
  col = line_colors,
  xlab = "Days",
  lty = 1,
  cex = 1.2,
  lwd = 2,
  font.lab = 2
)
legend(
  "topleft",
  inset = c(0.05, 0.05),
  legend = transitions,
  col = line_colors,
  lty = 1,
  lwd = 2,
  cex = 1.2,
  bty = "o"
)
```
